The Galactic Chemical Evolution of Carbon: Implications for Stellar Nucleosynthesis

### 2025-01-16 carbon_paper/tex/main.tex
updated draft with mcmc models

\newcommand{\Yct}{{y_{\rm C}}}
\newcommand{\Ycc}{{y_{\rm C}^{\rm CC}}}
\newcommand{\Yoc}{{y_{\rm Mg}^{\rm CC}}}
\newcommand{\Ycagb}{{y_{\rm C}^{\rm AGB}}}
\newcommand{\aagb}{\alpha_{\rm C}^{\rm AGB}}
\newcommand{\zagb}{\zeta_{\rm C}^{\rm AGB}}
\newcommand{\zcc}{\zeta_{\rm C}^{\rm CC}}
 \newcommand{\yb}{\ensuremath{\rotatebox[origin=B,y=0.5ex]{180}{y}}}
\newcommand{\y}{Y}
\newcommand{\fagb}{f_{\rm C}^{\rm AGB}}

@@

\newcommand{\dbstrike}[1]{{\color{Thistle} \sout{#1}}}
\newcommand{\dbadd}[1]{{\color{Thistle} #1}}
\newcommand{\dbnote}[1]{ {\color{Thistle} \textit{\small (DB: #1)}} }

@@

% Abstract of the paper
\begin{abstract}
% context
The origin of C, despite its importance, remains poorly understood. 
% 
We aim to constrain the stellar yields of C through multi-zone Galactic chemical evolution models by comparing predictions with \apogee\ subgiants abundances.
% 
We find that \caafe\ trends, when restricted in metallicity, are an empirical estimate of the delayed C sources, enabling us to estimate that \agb\ stars and \cc\ produce about 20\% and 80\% of C, respectively.  
The \caah\ trend instead represents the equilibrium abundances of C and Mg. 
We use the \caah\ trend to estimate the CCSNe C/Mg yield, determining that  $\Ycc/\Yoc = EQUATION$ when including \agb\ C. 
% misc
Our models are relatively independent of uniform scaling of yields and outflows, and alternate star formation histories. 
However, the stars which contribute to \agb\ C production and the \ia\ delay time distribution of Fe contribute uncertainties to our conclusions. 
% gas phase
While reliable gas-phase and low-metallicity measurements of C are challenging, we find that our model and a single-zone model with our recommended yields replicate the broad trends of \caah{} across different environments and metallicities.

@@

\begin{figure*}
    \centering
    \includegraphics{subgiants.pdf}
    \caption{The [C/Mg] ratio against [Mg/H] (left) and [Mg/Fe] (right) for the \citet{jack}~sample of \apogee{} subgiants. On the left, we plot high and low-$\alpha$ stars in blue and orange, using the separation defined in Equation \ref{eq:high_alpha} \dbstrike{(the high and low-$\alpha$ stars are named for their high or low $\alpha$-element (Mg) to Fe ratios)}. On the right, we colour-code stars according to their [Mg/H] abundance.} \label{fig:subgiants}
\end{figure*}

Despite being the most abundant metal in the Universe~\citep[e.g.][]{magg+22}, the nucleosynthetic origin of C is inadequately understood. \dbnote{need synonyms for poorly understood}
Broadly, studies agree that both high mass and asymptotic giant branch (AGB) stars produce C, but their relative yields are disputed.
From a Galactic chemical evolution (GCE) perspective, some authors have argued that high-mass stars should dominate~\citep[e.g.][]{prantzos+94, HEK00, romano+20, franchini+20, gustafsson22}, and others have argued that AGB stars should contribute at least half of the C in the universe~\citep[e.g.][]{tinsley79, chiappini+03, mattsson10, KKU11, rybizki+17, KKL20}. 
\dbadd{We aim to empirically characterize the yields of C and the ratio of AGB and CCSNe production.}

% Formed in the cores of stars during He fusion, C is the first \dbadd{primary? element} after He. Additionally, C is one of the only light elements formed in low-mass stars \citep[e.g.,][]{KL14}. \dbnote{I don't like how many e.g.'s there are, and is this the right citation here?}
%\dbadd{
% Because the effects of first dredge up are mass-dependent, C and N abundances are frequently used as age indicators for RGB stars \citep{MG15, martig16, hasselquist19, vincenzo+21}. 
% }
% \dbnote{C abundnces for stellar structure (opacities) and star formation (line cooling)}

Predictions of C yields from stellar evolution and supernova (SN) models are riddled with uncertainties (see, e.g., the reviews/summaries by \citealt{romano+10, KL14}).
Stellar yields are shaped by many poorly understood processes, such as mass loss rates \citep{sukhbold+16, beasor+2020}, rotational mixing \citep{frischknecht+16, LC18}, reaction rates \dbadd{citations}, and convection \citep{chieffi2001, ventura+13}.
\dbnote{I checked these, is sukhbold the best reference on mass loss?}
\dbadd{In addition, stellar parameters such as convection and mass loss are calibrated to match observed stars and cannot be predicted from first principles.} \dbnote{need to check if this means SE models are calibrated to match yields or just stellar parameters.}
Small changes in the initial mass of only $\sim$$0.01 M_\odot$  were recently shown to have dramatic effects on the evolution and explosion outcomes of massive stars \citep{bruenn+2023, vartanyan_burrows2023}.
Furthermore, many stars are found in binary systems \dbadd{(e.g. \citealt{MKB2019}, others?)}, the effects of which are almost entirely unexplored in stellar nucleosynthesis models \citet{farmer+21}.
Therefore, accurate yield predictions from stellar theory requires both a much greater understanding of stellar physics and a much more densely sampled grid of models than is currently available.

In light of the challenges of stellar theory, recent work has instead sought to characterize chemical yields empirically.
\citet{HLA20} measured the abundance ratios in the ejecta of Cassiopeia A using spatial resolved X-ray spectra.
\citet{weinberg+19, weinberg+22} analyzed the abundance ratio trends of high- and low-alpha sequence stars in APOGEE~\citep{apogee17} to infer relative yields of prompt and delayed enrichment sources and the metallicity dependence thereof.
\citet{emily+19, emily+22, emily+24} applied the same methodology to GALAH~\citep{DeSilva2015, Martell2017}.
\citet{rodriguez+21, rodriguez+23} measured the mean Fe yield of core-collapse supernovae (CCSNe) using the radioactive tails of their lightcurves.
\dbnote{even papers as far back as at least https://ui.adsabs.harvard.edu/abs/1997NuPhA.621...27G/abstract do this. Include this series as well: https://ui.adsabs.harvard.edu/abs/2023MNRAS.525.1329L/abstract}

% \note{Reading through at least at this moment, I'd say this is a reasonable place for a new paragraph. I wouldn't worry about referencing Weller et al. (2024) since He chemical evolution is not super relevant here.}
 Folowing many previous works  (e.g. \citealt{DTS78, BF06, prantzos+18, berg+19}; see also the recent reviews by \citealt{romano22} and \citealt{RM21}.) this paper focuses on the evolution of C in GCE models.
Most notably for our purposes, \citet{james+23} recently showed that the enrichment history of the Galactic disk can be described as an equilibrium phenomenon.
In equilibrium GCE, trends in abundance ratios with metallicity arise as a consequence of trends in yield ratios (see discussion in their section {\color{red} X.Y}).
\citet{james+23} had previously shown that the correlation between gas-phase N and O abundances (see their Fig. 1) could be described as a consequence of the metallicity dependence of the relative N and O yields.
The equilibrium scenario suggests that this argument should extend to any element whose enrichment by stellar populations does not require long timescales ($\gtrsim$few Gyr).
We therefore extend \citepos{james+23} arguments to C, which is closely related to N.

We use a sample of subgiant stars from APOGEE \citep{apogee17} as our primary observational constraint.
\dbadd{APOGEE Subgiants provide a large, homogenous sample of well-measured C abundances unaffected by stellar evolution}.
Subgiants advantageously have photospheric abundances directly representing birth C abundances \citep{gilroy89, korn+07, lind+08, souto+18, souto19}.
In other evolutionary phases, the measured abundances are known to be affected by internal processes (see discussion in section 2 below), namely gravitational settling in main sequence stars {\color{red} (reference)} and dredge-up in red giants {\color{red} (reference)}, 
Our sample should greatly mitigate these difficulties with inferring birth abundances of C, which have been a significant source of uncertainty for previous investigations of its nucleosynthetic origin.

@@

In Section~\ref{sec:data_selection} we describe the selection for our subgiant observational sample.
Section~\ref{sec:nucleosynthesis} discusses theoretical estimates of \agb\ and \cc\ C yields. We parameterize our yield choices and derive a estimate of C/Mg yields observationally.
Section~\ref{sec:vice} reviews the multizone model and modifications thereof for this work.
Section~\ref{sec:results} presents our multizone model predictions and GCE parameters which are observationally permissable. 
Finally Section~\ref{sec:gas} compares our model to literature collected observations of halo stars and extragalactic HII regions.

@@

% Subgiants are a useful empirical benchmark because their photospheric abundances should reflect their birth mixture. 
As an emperical benchmark, we use subgiant stars, whose photospheres accurately represent birth abundances. 
During main sequence evolution, metals can fall to below the convective envelope (i.e., ``gravitational settling").
When stars leave the main sequence and become subgiants, these metals are mixed with the convective envelope \citep[]{gratton+00, souto19}. 
 \textit{First dredge up} during the red giant phase brings CNO processed core material to the surface \citep{iben67, KL14}.

We use the sample of \apogee\ DR17 subgiants from \citet{jack}, selecting stars from allStar-dr17-synspec\_rev1 within the $\log g$-$T_\text{eff}$ polygon given by
 \begin{equation}
    \begin{cases} \label{eq:subgiant_selection}
        \log \text{g} \geq 3.5 \\
        \log \text{g} \leq 0.004\,T_{\rm eff} - 15.7 \\
        \log \text{g} \leq 0.0007\,T_{\rm eff} + 0.36 \\
        \log \text{g} \leq -0.0015\,T_{\rm eff} + 12.05 \\
        \log \text{g} \geq 0.0012\,T_{\rm eff} - 2.8. \\
    \end{cases}
\end{equation}
Also following \citet{jack}, we exclude stars with the flags:
        \verb|APOGEE_MIRCLUSTER_STAR|,
        \verb|APOGEE_EMISSION_STAR|,
        \verb|APOGEE_EMBEDDEDCLUSTER_STAR|,
        \verb|APOGEE2_YOUNG_CLUSTER|,
        \verb|APOGEE2_W345|,
        \verb|APOGEE2_EB|, and
        \verb|DUPLICATE|.
We additionally remove stars with no reported C, Mg, or Fe abundances (in the {\tt X\_FE} columns). These cuts yield a sample containing \nsubgiants\ subgiants.

Fig.~\ref{fig:subgiants} shows the subgiant sample plotted in \caah\ and \caafe{}.\footnotemark{} In this sample, [C/Mg] increases with metallicity, and [C/Mg] decreases with [Mg/Fe] at fixed [Mg/H].

\dbadd{
Not shown here, we investigate the median trends for different samples and surveys, including APOGEE Giants with mixing corrections \citep{vincenzo+21}, APOGEE dwarfs, Gaia-ESO dwarfs, GALAH subgiants, LAMOST MRS (CITE), and smaller spectroscopic samples of stars. The predominant difference between any survey with our sample is a systematic offset in [C/Mg] in both [Mg/Fe] and [Fe/H] of at most 0.2 dex. The exception is GALAH, who's C abundances are highly incomplete, based on a single line and thus C measurements are likely unreliable (see e.g. Emily). If the median trend of [C/Mg] is offset from what we assume, then we can simply scale our C yields by the appropriate factor (e.g. -0.15 dex => $10^{-0.15} \approx 0.7$, so $y_{\rm C} \rightarrow  0.7 y_{\rm C}$) without significantly changing our results.
}

@@

Table \ref{tab:fiducial_mod} contains our fiducial yields and abundance scale. We use solar abundances from \citet{magg+22} with a +0.04 gravitational settling correction  and assuming $X_\odot= 0.71, \mu_{H} = 1$ as in \citet{david_fe}. We thus assume $Z_\odot=0.016$ as apposed to $Z=0.014$ in \citet{asplund+09}.
% Z = (1-Y) / (1 + X/Z)

\begin{table}
	\centering
    \caption[]{Solar abundance scale and fiducial yields (in units of SSP~birth mass). See section \ref{sec:agb} for the definition of \fruity. The solar abundance scale is \citet{magg+22} + 0.04. }
	\label{tab:fiducial_mod}

@@

We define a stellar yield as the fraction of a star's initial mass $M$ which is converted into freshly synthesized C. If $\Delta Z_{\rm C}$ represents the change in the mass fraction of C abundance, then the stellar yield is
\begin{equation}
    \y_{\rm C}^{\rm AGB} =  \Delta Z_{\rm C} \frac{M_{\rm ejected}}{M} = 
    \frac{W_{\rm C}(M) - Z_{\rm C}\, M_{\rm ejected}}{M}
\end{equation}
where $\Delta Z_{\rm C}$ represents the change in mean metallicity of the initial composition the star was born from, $M_{\rm ejected} = M - M_{\rm rem}$ is the total ejected wind mass through the stars lifetime, $W_{\rm C}$ is the total mass of C ejected through winds, and $Z_{\rm C}$ is the initial C mass fraction of the star. 
For CCSNe, 
\begin{equation} \label{eq:imf-cc-yield}
   Y_{\rm C}^{\rm CC}(Z) = 
    \frac{E(M)\,M_{\rm C}(M) + W_{\rm C}(M) - Z_{\rm C}\,M}{M}
\end{equation}
\dbnote{Is the - Z M term right or should M be Mzams - Mrem ?}
where $E(M)$ is the explodability function, $M_{\rm C}(M)$ is the mass ejected in a successful supernovae, and $W_{\rm C}$ and $Z_{\rm C}$ are as above.
We assume 8\,$\Mo$ is the minimum mass of \cc\ progeneters (see \VICE\ documentation and \citealt{emily+21}). 
Note that a yield may be negative if a star expels gas with lower $Z_{\rm C}$ than the star was born with.

An IMF-averaged yield adds together the yields of stars of each mass, weighted by the fraction of stars of each mass (the IMF). 
The IMF-averaged yield $y_X^{\rm proc}$ for element $X$ under the enrichment process ${\rm proc}$ is given by
\begin{equation} \label{eq:imf-agb-yield}
    y_{X}^{\rm proc}(Z,\tau) = 
    \frac{
    \int_{M_{\rm to}(\tau)}^{M_{\rm max}} 
    \y_{\rm X}^{\rm proc}(M, Z) M\ 
    \frac{dN}{dM}\ dM
}
{
    \int_{M_{\rm min}}^{M_{\rm max}}\ M\ \frac{dN}{dM}\ dM
}
\end{equation}
where ${dN}/{dM}$ is the IMF and $M_{\rm to}(\tau)$ is the mass of stars with lifetime $\tau$, and $M_{\rm min}=0.08$ and $M_{\rm max} = 100$ are the assumed maximum and minimum values of the IMF.
$y$ thus represents the amount of freshly synthesized material returned to the ISM per unit of star formation. 
Note that in detail, the core collapse enrichment occurs on such short timescales that $M_{\rm to} = M_{\rm min, cc} = 8\,\Mo$

@@

\begin{figure*}
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[]{\dbnote{replace y label with stellar C yield}. The net fractional \agb\ C yield  plotted as a function of initial stellar mass $M$ and colour-coded according to metallicity. The black dashed line shows $\y=0$ for reference. Each panel represents yields from one of four \agb\ models: \fruity{}, \aton{}, \monash{}, \nugrid{} (see sections \ref{sec:agb}) }

@@

    \caption[]{\textbf{Left} \dbnote{ylabel: integrated carbon AGB yield} The (IMF-weighted) \agb\ C yield $\Ycagb$ as a function of metallicity for each of the \agb\ yield models. ($\Ycagb$ is the net mass of C produced by \agb\ stars per unit mass of star formation, after 10\,Gyr and assuming a \citealt{kroupa01} IMF.)
    \textbf{Right} cumulative C return as a function of age for a solar metallicity single stellar population. The dashed black line shows the delay time distribution of type Ia supernovae ($\propto t^{-1.1}$) for comparison. The minimum of \aton{} is $y(t)/y(t_{\rm end})=-46$. \dbnote{I think this changed with the new integration scheme...}
}

@@

Both HBB and TDU result in mass and metallicity dependent C yields. 
Stars less than \about{2}\,$\Mo$ do not experience TDU As a result, these stars C abundances are only affected by first dredge-up (CITE), resulting in small net C yields.
Above \about{2}\,$\Mo$, TDU becomes important, enriching the outer layers with C.
In \agb\ stars more massive than \about{5}\,$\Mo$, efficient HBB turns most \C[12] into $^{14}$N.
Metal poor stars dredge up more material \dbnote{for which physics, compactness, mixing efficiency( Boothroyd \& Sackmann 1988))? etc.}, resulting in elevated C production \citep[e.g.][]{ventura+13}.
This is due to  the more compact and higher core masses in low-mass stars, causing a stronger thin-shell instability leading to stronger pulse-driven convections and dredging up more C. Note that HBB also increases with lower metallicity, but the effects of stronger TDU outweighs this. Additionally, as stars evolve through the TPAGB, they dredge up more material for the same reasons. The number of pulses is driven by the chosen mass loss formulation.

>>>>

\newcommand{\Yct}{{y_{\rm C}}}
\newcommand{\Ycc}{{y_{\rm C}^{\rm CC}}}
\newcommand{\Yoc}{{y_{\rm Mg}^{\rm CC}}}
\newcommand{\Ycagb}{{y_{\rm C}^{\rm AGB}}}
\newcommand{\ycagb}{\y_{\rm C}^{\rm AGB}}
\newcommand{\aagb}{\alpha_{\rm C}^{\rm AGB}}
\newcommand{\zagb}{\zeta_{\rm C}^{\rm AGB}}
\newcommand{\zcc}{\zeta_{\rm C}^{\rm CC}}
\newcommand{\noneqagb}{\nu_{\rm C}^{\rm AGB}}
 \newcommand{\yb}{\ensuremath{\rotatebox[origin=B,y=0.5ex]{180}{y}}}
\newcommand{\y}{Y}
\newcommand{\fagb}{f_{\rm C}^{\rm AGB}}

\newcommand{\zetao}{\zeta^{(0)}}
\newcommand{\zetai}{\zeta^{(1)}}
\newcommand{\zetaii}{\zeta^{(2)}}

@@

\newcommand{\dbstrike}[1]{{\color{Thistle} \sout{#1} }}
\newcommand{\dbadd}[1]{{\color{Thistle} #1}}
\newcommand{\dbnote}[1]{ {\color{Thistle} \textit{\small (DB: #1)}} }

@@

% Abstract of the paper
\begin{abstract}
% context
The astrophysical origin of carbon remains largely uncertain---different studies propose that either asymptotic giant branch (AGB) stars or core collapse supernovae (CCSNe) may be the dominant process.
% 
We aim to constrain the stellar yields of C through multi-zone Galactic chemical evolution models by comparing predictions with \apogee\ subgiants abundances.
% 
We find that \caafe\ trends, when restricted in metallicity, are an empirical estimate of the delayed C sources, enabling us to estimate that \agb\ stars and \cc\ produce about 20\% and 80\% of C, respectively.  
The \caah\ trend instead represents the equilibrium abundances of C and Mg. 
We use the \caah\ trend to estimate total C/Mg yield ratio, determining that  $\Ycc/\Yoc = EQUATION$.
% misc
Our models are relatively independent of uniform scaling of yields and outflows, and alternate star formation histories. 
However, the stars which contribute to \agb\ C production and the \ia\ delay time distribution of Fe contribute uncertainties to our conclusions. 
% gas phase
While reliable gas-phase and low-metallicity measurements of C are challenging, we find that our model and a single-zone model with our recommended yields replicate the broad trends of \caah{} across different environments and metallicities.

@@

\begin{figure*}
    \centering
    \includegraphics{subgiants.pdf}
    \caption{The [C/Mg] ratio against [Mg/H] (left) and [Mg/Fe] (right) for the \citet{jack}~sample of \apogee{} subgiants. On the left, we plot high and low-$\alpha$ stars in blue and orange, using the separation defined in Eq.~\ref{eq:high_alpha}.  On the right, we colour-code stars according to their [Mg/H] abundance.} \label{fig:subgiants}
\end{figure*}

Despite being the most abundant metal in the Universe~\citep[e.g.][]{magg+22}, the nucleosynthetic origin of C is poorly understood. 
There is broad agreement that both high mass and asymptotic giant branch (AGB) stars produce C \citep{jennifer19}, but their relative yields are disputed.
From a Galactic chemical evolution (GCE) perspective, some authors have argued that high-mass stars should dominate~\citep[e.g.][]{prantzos+94, HEK00, romano+20, franchini+20, gustafsson22}, and others have argued that AGB stars should contribute at least half of the C in the universe~\citep[e.g.][]{tinsley79, chiappini+03, mattsson10, KKU11, rybizki+17, KKL20}. 
In this paper, we aim to constrain C production using subgiant stars and multi-zone GCE models \dbstrike{incorporating stellar radial migration}.

% Formed in the cores of stars during He fusion, C is the first \dbadd{primary? element} after He. Additionally, C is one of the only light elements formed in low-mass stars \citep[e.g.,][]{KL14}. \dbnote{I don't like how many e.g.'s there are, and is this the right citation here?}
%\dbadd{
% Because the effects of first dredge up are mass-dependent, C and N abundances are frequently used as age indicators for RGB stars \citep{MG15, martig16, hasselquist19, vincenzo+21}. 
% }
% \dbnote{C abundnces for stellar structure (opacities) and star formation (line cooling)}
% \dbadd{In addition, stellar parameters such as convection and mass loss are calibrated to match observed stars and cannot be predicted from first principles.} \dbnote{need to check if this means SE models are calibrated to match yields or just stellar parameters.}

Predictions of C yields from stellar evolution and supernova (SN) models are riddled with uncertainties (see, e.g., the reviews by \citealt{romano+10, KL14}).
Some \dbadd{substantial uncertanties} include mass loss rates \citep{sukhbold+16, beasor+2020}, rotational mixing \citep{frischknecht+16, LC18}, nuclear reaction rates \dbadd{CITE}, and convection \citep{chieffi2001, ventura+13}.
Recent work also suggests that small changes in a stars's initial mass of 
 $\sim$$0.01 M_\odot$  have significant effects on the evolution and explosion outcomes of massive stars \citep{bruenn+2023, vartanyan_burrows2023}.
Furthermore, many \add{massive} stars are found in binary systems \dbadd{\url{https://ui.adsabs.harvard.edu/abs/2012Sci...337..444S/abstract} ??}, the effects of which are almost entirely
unexplored in stellar nucleosynthesis models \citet{farmer+21}.
Accurate predictions of metal yields from stellar evolution models remain elusive.

Given the challenge of {\it a priori} yield predictions, recent work aims to empirically characterize elemental production \dbnote{does it make sense to frame this as only recent work? I just don't know but I feel like there are very early examples which at least use observations as a guide for yields, like maybe even B2FH. or garnett1997}
\citet{weinberg+19, weinberg+22} analyzed the abundance ratio trends of high- and low-alpha sequence stars in APOGEE~\citep{apogee17} to infer relative yields of prompt and delayed enrichment sources and the metallicity dependence thereof.
\citet{emily+19, emily+22, emily+24} applied the same methodology to GALAH~\citep{DeSilva2015, Martell2017}.
\citet{rodriguez+21, rodriguez+23} measured the mean Fe yield of core-collapse supernovae (CCSNe) using the radioactive tails of their lightcurves.
Their measurements, provide one of few empirical anchors on the absolute scale of yields \citep{david_fe}.
\dbadd{Additionally the series of papers including \citet{eitner+2020} includes some relevant work.}

 % Folowing many previous works  (e.g. \citealt{DTS78, BF06, prantzos+18, berg+19}; see also the recent reviews by \citealt{romano22} and \citealt{RM21}.) this paper focuses on the evolution of C in GCE models.
\citet{james+24} recently argued that the enrichment history of the Galactic disk can be described as an equilibrium phenomenon.
In equilibrium GCE, trends in abundance ratios closely trace yield ratios (see their section 5.2.2).
\citet{james+23} had previously shown that the correlation between gas-phase N and O abundances (see their Fig. 1) could be described as a consequence of the metallicity dependence of the relative N and O yields.
Under their equilibrium scenario, this argument should extend to any element whose enrichment by stellar populations does not require long timescales ($\gtrsim$few Gyr).
We show in \add{REF 3.1} that the characteristic delay times of C production should be $\sim$ few hundred Myr, within this limit. 
\dbnote{This is complicated by the difference between high alpha and low alpha, and 1-2Msun stars do produce carbon...}

We use a sample of subgiant stars from APOGEE \citep{apogee17} as our primary observational constraint.
APOGEE Subgiants provide a large, homogeneous sample of well-measured C abundances unaffected by stellar evolution.
\dbadd{Subgiants' photospheric C abundance closely matches their birth composition.}
\dbstrike{Subgiants have the advantage that their photospheric abundances of C directly reflect their birth C abundances much more closely than other spectral types.} \citep{gilroy89, korn+07, lind+08, souto+18, souto19}.
In other evolutionary phases, the measured abundances are known to be affected by internal processes (see discussion in section 2 below).
Our sample should therefore be unaffected by uncertainties with inferring their birth abundances.

@@

In Section~\ref{sec:data_selection} we describe the selection for our subgiant observational sample.
In Section~\ref{sec:nucleosynthesis}, we discusses theoretical estimates of \agb\ and \cc\ C yields and present the yield prescriptions we use for our models.
In Section~\ref{sec:vice}, we describe the remaining details of our GCE models.
In Section~\ref{sec:results}, we present our model predictions and discuss their observationally distinguishable predictions. 
In Section~\ref{sec:gas}, we compare our model to observations of halo stars and extragalactic HII regions from the literature. 
We summarize our conclusins in section~\ref{sec:conclusions}.

@@

% Subgiants are a useful empirical benchmark because their photospheric abundances should reflect their birth mixture. 
We use subgiant stars as our empirical benchmark.
Subgiants have the advantage that their photospheric C and N abundances should closely approximate their birth mixture.
During main sequence evolution, metals can fall below the convective envelope (i.e., ``gravitational settling''; e.g. \dbadd{CITE}).
When stars evolve off main sequence, these metals are reincorporated into the convective envelope \citep[]{gratton+00, souto19}. 
When a star becomes a red giant, \textit{first dredge up} pollutes its photosphere with core, CNO-processed material
\citep{iben67, KL14}.

We use the sample of \apogee\ DR17 subgiants from \citet{jack}. We select stars from the ASPCAP pipeline (revision 1) within the $\log g$-$T_\text{eff}$ polygon given by
 \begin{equation}
    \begin{cases} \label{eq:subgiant_selection}
        \log \text{g} \geq 3.5 \\
        \log \text{g} \leq 0.004\,T_{\rm eff} - 15.7 \\
        \log \text{g} \leq 0.0007\,T_{\rm eff} + 0.36 \\
        \log \text{g} \leq -0.0015\,T_{\rm eff} + 12.05 \\
        \log \text{g} \geq 0.0012\,T_{\rm eff} - 2.8. \\
    \end{cases}
\end{equation}
Following \citet{jack}, we also exclude stars with the flags:
        \verb|APOGEE_MIRCLUSTER_STAR|,
        \verb|APOGEE_EMISSION_STAR|,
        \verb|APOGEE_EMBEDDEDCLUSTER_STAR|,
        \verb|APOGEE2_YOUNG_CLUSTER|,
        \verb|APOGEE2_W345|,
        \verb|APOGEE2_EB|, and
        \verb|DUPLICATE|.
We additionally remove stars with no reported (or flagged unreliable), C, Mg, or Fe abundances. These cuts yield a sample containing \nsubgiants\ subgiants.

Fig.~\ref{fig:subgiants} shows out subgiant sample plotted in the \caah\ and \caafe{} planes.\footnotemark{} In this sample, [C/Mg] increases with [Mg/H], and [C/Mg] decreases with [Mg/Fe] at fixed [Mg/H].

We have also investigated these thrends in other catlogues (not show) in order to understand the systematic uncertainties of our APOGEE subgiant sample.
Namely, we we consider subgiants/dwarfs from GALAH DR4, Gaia-ESO DR5.1, and LAMOST with similar selection criteria. Additionally, we consider APOGEE Giants with first dredge up mixing corrections from \citep{vincenzo+21}.
In general, different surveys follow similar trends in \caah\ and \caafe\ but with systematic offsets at the $\sim{0.1}$ dex level in [C/Mg] given [Mg/H] and slight variations in the slope.
\dbnote{Galah DR4 fixes the issues we were seeing and updates how they process C.}
These systematic differences are not a major concern for our conclusions since our models are mostly constrained by the slope of \caah\ and \caafe,  not their normalization (see Fig.~7? and discussion in section \dbadd{5.2}).

@@

Following \citet{james+23}, we integrate yields and our GCE models using the publicly available Versatile Integrator for Chemical Evolution (\VICE, \citealt{JW20}). We retain their prescription for O, Fe, an yields, scaled down to the normalization recommended by \citet{david_fe} based on .....
Table~\ref{tab:fiducial_mod} summarizes these choices. As required by \VICE, these yields quantify the mass of some element that is newly produced (i.e. net yields as opposed to gross yields). The SN yields are in units of the proginitors initial mass, while the AGB yields are defined relative to the star's zero-age main sequence mass.

\dbadd{paragraph about yields in general}
A yield quantifies the new production of a chemical species. For a given star, we define the {\it net fractional yield} to be the mean change in the chemical abundance from the birth to ejected mass, multiplied by the ejected fraction of a stars mass, i.e.
\begin{equation} 
   Y_{\rm C}^{\rm AGB}(Z) = 
    \delta Z_X \frac{M_{\rm ej}}{M}
\end{equation}
Note that yields may become negative if more of the initial chemical abundance of X is transformed into other elements than other elements transformed into X.
In general, we describe yields by their population-averaged forms, the amount of newly produced X per unit mass of star formation (from supernovae and stellar winds).
\begin{equation} 
   y_{X}(Z) = 
    \frac{M_X - M_{\rm X, ini}}{M_{\rm ssp}}
\end{equation}

In \VICE, massive star nucleosynthesis is approximated as occurring instantaneously after a star formation event. Under this approximation, the yield is simply a constant  of proportionality between the metal production rate and the SFR:
\begin{equation}
    \dot{M}_X^{\rm CC} = y_{X}^{\rm cc}\, \dot{M}_\star
\end{equation}
The population-averaged yield, $y_{X}^{\rm cc}$ can be expressed as
\begin{equation}
y_{X}^{\rm CC} = \frac{
\int_8^u (E(m) m_X + w_X - Z_X(m - m_{\rm rem}) )\frac{dN}{dM}\ dM
}{
\int_l^u m\,\frac{dN}{dM}\ dm
}
\end{equation}
where $m_X$ and $w_X$ are the mass of X ejected in a CCSN explosion and through stellar winds respectively, from a star of initial mass $m$. $E((m)$ is a function describing the fraction of stars of mass $m$ that successfully explode (se discussion in \S \dbadd{3.2} below). $m_{\rm rem}$ is the mass of the remnant left behind, $dN/dM$ is the IMF, for which we assume \dbadd{cite}. $l$ and $u$ are the lower and upper mass limits of star formation. $8\,\Mo$ is the minimum mass of a CCSN progenitor. In section \S 3.2 below, we compute the values of $\Ycc$ using \VICE's `vice.yields.ccsne.fractional` function and discuss theoretical expectations of C yields based on massive star evolutionary models.

The population averaged SNe Ia yield, $y_{X}^{\rm ia}$, is given by,
\begin{equation}
    y_{X}^{\rm Ia} = m_{x}^{\rm Ia} \frac{N_{\rm Ia}}{m_\star}
\end{equation}
where $m_{X}^{\rm Ia}$ is the average mass of X produced in a single SN Ia event, and $N_{\rm Ia} / M_\star$ is the number of such events per unit mass of star formation. The latter can be expressed as an integral over the SN Ia delay-time-distribution (DTD), which describes the rate at which stellar populations produce X through SN Ia in \VICE. Following \citet{james+23}, we use a $t^{-1.1}$ single power-law with a minimum delay-time of 150 Myr based on comparisons of the SN IA rate as a function of redshift and the cosmic SFR (e.g. \citealt{maoz+12}).

The net yield from a single AGB star is given by \begin{equation} \label{eq:net-yield}
   Y_{\rm C}^{\rm AGB}(Z) = 
    \delta Z_X \frac{M_{\rm ej}}{M}
\end{equation}
where $M_{\rm ej}$ is the mass of the ejected envelope, and $\delta Z_X$ is the average change in the abundance of X in the envelope over the lifetime of the star. Models of AGB star nucleosynthesis report values of $Y_X^{\rm AGB}$ sampled on a grid of initial masses and metallicities (see discussion in \S \dbadd{3.1}), and \VICE\ accepts these tables directly as input. The population-averaged AGB star yield is then given by 
\begin{equation} \label{eq:imf-agb-yield}
    y_{X}^{\rm agb}(Z,\tau) = 
    \frac{
    \int_{l}^{M_{\rm max}^{\rm AGB}} 
    \y_{\rm X}^{\rm proc}(M, Z) M\ 
    \frac{dN}{dM}\ dM
}
{
    \int_{l}^{u}\ M\ \frac{dN}{dM}\ dM
}
\end{equation}
where $m_{\rm min}$ is the minimum and maximum initial mass of stars that become AGBs. While $m_{\rm min}$ theoretically extends as low as $0.1 \Mo$, we interpolate yields to zero past $0.5 \Mo$ as these stars live longer than the disk and no yield grid extends this low. In Section~\ref{sec:agb},  we discuss theoretical expectations for C production in AGB stars and our adopted tables of $\Ycagb$ from the literature.
\dbadd{specify exactly how we change the yield interpolation for AGB stars}

\begin{table}
	\centering
    \caption[]{Solar abundance scale and fiducial yields (in units of stellar population birth mass). See section \ref{sec:agb} for the definition of \fruity. The solar abundance scale is \citet{magg+22} + 0.04. }
	\label{tab:fiducial_mod}

@@

@@

\begin{figure*}
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[]{\dbnote{replace y label with stellar C yield, draw actual interpolations...}. The net fractional \agb\ C yield  plotted as a function of initial stellar mass $M$ and colour-coded according to metallicity. The black dashed line shows $\y=0$ for reference. Each panel represents yields from one of four \agb\ models: \fruity{}, \aton{}, \monash{}, \nugrid{} (see sections \ref{sec:agb}) }

@@

    \caption[]{\textbf{Left} \dbnote{ylabel: integrated carbon AGB yield} The (IMF-weighted) \agb\ C yield $\Ycagb$ as a function of metallicity for each of the \agb\ yield models. ($\Ycagb$ is the net mass of C produced by \agb\ stars per unit mass of star formation, after 10\,Gyr and assuming a \citealt{kroupa01} IMF.)
    \textbf{Right} cumulative C return as a function of age for a solar metallicity single stellar population. The dashed black line shows the delay time distribution of type Ia supernovae ($\propto t^{-1.1}$) for comparison. The minimum of \aton{} is at $y( 0.30\,{\rm Gyr})/y(t_{\rm end})=-7.81$ and is large as the net yield is close to zero.
}

@@

Both HBB and TDU result in mass- and metallicity-dependent C yields. 
Stars less than \about{2}\,$\Mo$ do not experience TDU. As a result, C yields from these stars are affected only by first dredge-up \citep[Table 1]{karakas10}, resulting in small net C yields.
Above \about{2}\,$\Mo$, TDU becomes important, enriching the outer layers with C.
In \agb\ stars more massive than \about{5}\,$\Mo$, efficient HBB turns most \C[12] into $^{14}$N.
TDU is more efficient for more compact, massive cores. HBB requires a more massive AGB star to reach sufficient temperatures at the base of the convective envelope to initiate proton captures. Both processes are more efficient at low metallicity due to the lower opacities and, more compact internal structures, and hotter temperatures. These two processes drive C yields from AGB stars, but their details are ultimantly set by the uncertain choices for stellar evolution models.

% compactness causes stronger thin-shell instability
% C yields increase because TDU outweights HBB when low Z.
